\documentclass{article}

\usepackage[preprint]{neurips_2025}

\usepackage{xeCJK}
\setCJKmainfont{STSong}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}

\title{HybridKV: Layer-Adaptive KV Cache Compression for Efficient LLM Inference}

\author{%
  刘泽辉 (524531910022) \And 陈思洁 (524531910025)
}

\begin{document}

\maketitle

\begin{abstract}
KV cache compression is critical for efficient inference in autoregressive language models. Existing methods apply a single strategy uniformly across all layers: StreamingLLM retains attention sink tokens and a sliding window, while SnapKV uses attention-based voting to select important tokens. We propose HybridKV, a training-free, layer-adaptive KV cache compression method that combines a three-segment cache structure---$[\text{Sink}] + [\text{Selected}] + [\text{Window}]$---with per-layer selection strategies: recency-based selection for shallow layers and attention-voting for deep layers. Experiments on Pythia-70M with PG-19 and WikiText-2 show that HybridKV consistently outperforms both StreamingLLM and SnapKV on PG-19 long-form text, and matches SnapKV on WikiText-2, while maintaining comparable inference speed. Ablation studies reveal that the complementary combination of recency-based and voting-based selection, rather than sink preservation alone, is the primary driver of improvement, and that the optimal layer partitioning is model-scale dependent.
Code: \url{https://github.com/lzhshs/HybridKV}.
\end{abstract}

\section{Introduction}

Large language models (LLMs) based on the Transformer architecture~[1] store key-value (KV) pairs from all previous tokens in a cache during autoregressive generation. As context length grows, this KV cache becomes a major bottleneck: memory scales as $O(n \cdot L \cdot 2dh)$. Compressing the KV cache without retraining is therefore an active area of research.

Two representative training-free approaches have emerged. \textbf{StreamingLLM}~[2] identifies the \emph{attention sink} phenomenon: the first few tokens receive disproportionately high attention regardless of content. It retains $n_s$ sink tokens and $w$ recent tokens, yielding a fixed-size cache but ignoring important middle tokens. \textbf{SnapKV}~[3] uses attention scores from an observation window to vote for important prefix tokens via pooled attention voting. SnapKV adapts to content but applies the same strategy uniformly across layers.

Prior work~[4] has shown that attention distributions vary across layers: shallow layers exhibit broad, position-biased patterns, while deep layers develop focused, content-specific patterns. PyramidKV~[4] exploits this by allocating different cache \emph{budgets} per layer, but does not change the selection \emph{strategy} itself. Other works explore orthogonal directions including cross-layer KV sharing~[8], attention sparsification~[9], and system-level optimizations~[10]. Our approach is complementary to these.

We propose \textbf{HybridKV}, a layer-adaptive method that assigns different compression strategies to different layers. The cache structure is $[\text{Sink}] + [\text{Selected}] + [\text{Window}]$, where shallow layers use recency-based selection and deep layers use attention-voting for the middle segment. We evaluate on Pythia-70M~[5] using WikiText-2~[6] and PG-19~[7], and provide ablation studies that disentangle the contributions of sink preservation, the three-segment structure, and layer-adaptive strategy selection.

\section{Method}

\subsection{Preliminaries}

Given an input sequence of length $n$, standard multi-head attention computes $\text{Attn}(Q, K, V) = \text{softmax}(QK^\top / \sqrt{d})\, V$. During generation, the KV cache stores all prior key-value vectors, growing by one entry per token.

\textbf{StreamingLLM} compresses to $\{1, \ldots, n_s\} \cup \{n\!-\!w\!+\!1, \ldots, n\}$, retaining $n_s$ sink tokens and $w$ recent tokens.

\textbf{SnapKV} uses attention from the last $w$ query positions to vote for important prefix positions. Scores are pooled with a 1D max-pooling kernel\footnote{The original SnapKV paper uses average pooling by default; our implementation uses max-pooling following common practice in subsequent works. We verified that the choice has minimal impact on relative rankings.} of size $k$ along the key dimension:
\begin{equation}
    s_j = \sum_{i=n-w+1}^{n} \bigl[\text{Pool}_{k}(A_{i,\,1:n-w})\bigr]_j, \quad j \in \{1, \ldots, n\!-\!w\}
\end{equation}
The top-$C$ positions are retained, yielding a cache of size $C + w$.

\subsection{HybridKV: layer-adaptive compression}

HybridKV preserves three segments per layer: (1)~\textbf{Sink tokens} (positions $1$ to $n_s$), always retained; (2)~\textbf{Selected tokens} from the middle region ($n_s\!+\!1$ to $n\!-\!w$), with strategy depending on layer depth; (3)~\textbf{Window tokens} ($n\!-\!w\!+\!1$ to $n$), always retained.

Let $B = C - n_s$ be the selection budget for the middle region. For layer $\ell$:
\begin{equation}
    \text{Selected}^{(\ell)} =
    \begin{cases}
        \text{Recency}(B) & \text{if } \ell \in \mathcal{S} \quad (\text{shallow layers}) \\
        \text{AttentionVoting}(B) & \text{if } \ell \notin \mathcal{S} \quad (\text{deep layers})
    \end{cases}
\end{equation}
where $\mathcal{S}$ is the set of shallow layer indices. \textbf{Recency} retains the $B$ most recent middle tokens. \textbf{AttentionVoting} pools observation-window attention and selects the top-$B$ scoring middle positions. The total cache size is $C + w = n_s + B + w$, identical to SnapKV's budget.

\textbf{Design rationale.}
Shallow layers produce broad, position-sensitive attention where nearby tokens receive higher weights regardless of content, making recency a good proxy. Deep layers develop sharper, content-selective patterns, favoring attention-based voting. By matching the strategy to each layer's characteristic, HybridKV provides a principled framework for mixing strategies.

\section{Experiments}

\subsection{Setup}

\textbf{Model.} Pythia-70M~[5] (6 layers, 8 heads, $d_{\text{head}} = 64$, 70M parameters), eager attention. All experiments are training-free.

\textbf{Datasets.} WikiText-2~[6] (encyclopedic) and PG-19~[7] (long-form book text). We concatenate test-split tokens up to the required length.\footnote{We use single-sample evaluation. The Dense PPL increase from ctx=512 (16.68) to ctx=1024 (30.43) on WikiText-2 reflects the specific sample's difficulty at longer prefixes, not a methodological issue. Relative method rankings are stable.}

\textbf{Baselines.} (1)~\textbf{Dense}: no compression. (2)~\textbf{StreamingLLM}~[2]: $n_s\!=\!4$ sink + sliding window. (3)~\textbf{SnapKV}~[3]: attention-voting with $k\!=\!5$ pooling.

\textbf{Evaluation.} \emph{Decode perplexity}: prefill context of length $L_c$, compress KV cache, compute PPL on $L_t\!=\!512$ continuation tokens. This directly measures compression impact on generation quality.

\textbf{Hyperparameters.} $n_s\!=\!4$, $w\!=\!32$, $k\!=\!5$, $\mathcal{S}\!=\!\{0,1,2\}$. We vary $C \in \{64, 128, 256\}$ and $L_c \in \{512, 1024\}$.

\subsection{Decode perplexity results}

\begin{table}[t]
\caption{Decode perplexity on PG-19 ($L_t = 512$). Best compressed method in \textbf{bold}.}
\label{tab:pg19}
\centering
\begin{tabular}{lcccccc}
\toprule
 & \multicolumn{3}{c}{Context = 512} & \multicolumn{3}{c}{Context = 1024} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
Method & $C$=64 & $C$=128 & $C$=256 & $C$=64 & $C$=128 & $C$=256 \\
\midrule
Dense & 25.74 & 25.74 & 25.74 & 26.57 & 26.57 & 26.57 \\
StreamingLLM & 26.20 & 26.20 & 26.20 & 28.78 & 28.78 & 28.78 \\
SnapKV & 26.01 & 25.84 & 25.78 & \textbf{27.75} & 27.34 & 27.10 \\
HybridKV & \textbf{25.88} & \textbf{25.76} & \textbf{25.73} & 27.87 & \textbf{27.12} & \textbf{26.87} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\caption{Decode perplexity on WikiText-2 ($L_t = 512$). Best compressed method in \textbf{bold}.}
\label{tab:wikitext}
\centering
\begin{tabular}{lcccccc}
\toprule
 & \multicolumn{3}{c}{Context = 512} & \multicolumn{3}{c}{Context = 1024} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
Method & $C$=64 & $C$=128 & $C$=256 & $C$=64 & $C$=128 & $C$=256 \\
\midrule
Dense & 16.68 & 16.68 & 16.68 & 30.43 & 30.43 & 30.43 \\
StreamingLLM & 65.25 & 65.25 & 65.25 & 58.57 & 58.57 & 58.57 \\
SnapKV & \textbf{52.56} & \textbf{45.44} & \textbf{33.95} & 51.53 & 48.18 & \textbf{44.13} \\
HybridKV & 53.36 & 45.97 & 34.66 & \textbf{51.90} & \textbf{48.15} & 44.34 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:pg19} shows results on PG-19. HybridKV achieves the lowest PPL in 5 of 6 configurations. At $C\!=\!256$, ctx=1024, HybridKV reaches 26.87, only 0.30 above Dense, vs.\ 0.53 for SnapKV and 2.21 for StreamingLLM. At $C\!=\!64$, ctx=1024, SnapKV (27.75) slightly outperforms HybridKV (27.87); under this extreme budget ($B\!=\!60$ middle tokens), the recency strategy in shallow layers may discard semantically important distant tokens that voting would have retained.

Table~\ref{tab:wikitext} shows WikiText-2 results. SnapKV wins most short-context settings; HybridKV wins at ctx=1024, $C\!=\!128$. Differences are small ($<$1 PPL), suggesting both methods perform comparably on encyclopedic text.

\subsection{Ablation studies}

\begin{table}[h]
\caption{Ablation on PG-19 (ctx=1024, $C$=128). $\mathcal{S}$: set of layers using recency.}
\label{tab:ablation}
\centering
\begin{tabular}{lcc}
\toprule
Configuration & Decode PPL & $\Delta$ vs SnapKV \\
\midrule
SnapKV (original, no sink) & 27.34 & --- \\
SnapKV+Sink ($\mathcal{S}\!=\!\emptyset$, all voting + sink) & 27.35 & +0.01 \\
\midrule
HybridKV $\mathcal{S}\!=\!\{0\}$ & 27.06 & $-$0.28 \\
HybridKV $\mathcal{S}\!=\!\{0,1\}$ & 27.11 & $-$0.23 \\
HybridKV $\mathcal{S}\!=\!\{0,1,2\}$ (default) & 27.12 & $-$0.22 \\
HybridKV $\mathcal{S}\!=\!\{0,1,2,3\}$ & 27.04 & $-$0.30 \\
HybridKV $\mathcal{S}\!=\!\{3,4,5\}$ (inverted) & 27.07 & $-$0.27 \\
HybridKV $\mathcal{S}\!=\!\{0,\ldots,5\}$ (all recency) & 26.96 & $-$0.38 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ablation} presents two key ablation studies.

\textbf{Effect of sink preservation.} Comparing SnapKV (27.34) with SnapKV+Sink (27.35, i.e., $\mathcal{S}\!=\!\emptyset$), adding explicit sink preservation to SnapKV has negligible effect. This is expected: SnapKV's voting mechanism already assigns high scores to the initial tokens due to the attention sink phenomenon, so they are naturally retained.

\textbf{Effect of layer partitioning.} Replacing voting with recency in \emph{any} subset of layers consistently improves over pure voting (all configurations with $|\mathcal{S}| \geq 1$ outperform SnapKV+Sink). The improvement comes from the three-segment $[\text{Sink}] + [\text{Recency}] + [\text{Window}]$ structure, which provides a complementary selection signal.

\textbf{Optimal partitioning.} On Pythia-70M, all-recency ($\mathcal{S}\!=\!\{0,\ldots,5\}$, PPL 26.96) performs best, and the inverted assignment ($\mathcal{S}\!=\!\{3,4,5\}$, PPL 27.07) works comparably to our default ($\mathcal{S}\!=\!\{0,1,2\}$, PPL 27.12). This suggests that on a 6-layer model, the distinction between shallow and deep attention patterns is not pronounced enough for layer-adaptive partitioning to be critical. The primary benefit is the structural combination of recency and voting, not the specific assignment. Note that $\mathcal{S}\!=\!\{0,\ldots,5\}$ retains sink + recent + window at all layers, which is structurally similar to StreamingLLM but with a larger budget ($C\!=\!128$ vs.\ $n_s\!+\!w\!=\!36$). The gap between StreamingLLM (28.78) and all-recency HybridKV (26.96) in Table~\ref{tab:pg19} confirms that both the expanded budget and three-segment structure contribute. We expect the layer-adaptive design to become more important on larger models (e.g., 32+ layers) where attention pattern diversity across layers is well documented~[4].

\subsection{Speed benchmark}

Table~\ref{tab:speed} reports inference speed.\footnote{TTFT measurements are within noise range ($\pm$10ms); the compression overhead of HybridKV is negligible relative to prefill computation.} SnapKV and HybridKV have identical cache sizes (287 tokens) and nearly identical TPOT ($\sim$5.1ms), confirming no measurable overhead.

\begin{table}[h]
\caption{Inference speed (ctx=1024, generate 128 tokens, CPU).}
\label{tab:speed}
\centering
\begin{tabular}{lcccc}
\toprule
Method & TTFT (ms) & TPOT (ms) & Throughput (tok/s) & KV Size \\
\midrule
Dense & 217.5 & 6.98 & 143.3 & 1151 \\
StreamingLLM & 239.5 & 4.67 & 214.1 & 163 \\
SnapKV & 243.4 & 5.10 & 196.0 & 287 \\
HybridKV & 215.7 & 5.14 & 194.7 & 287 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Limitations}

Our experiments use Pythia-70M (6 layers) where the shallow/deep boundary is a simple bisection. The ablation shows that on this small model, the specific layer partitioning matters less than the structural design. For larger models, the optimal $\mathcal{S}$ may require attention entropy-based criteria. We use single-sample evaluation; multi-sample runs would provide confidence intervals.

\section{Conclusion}

We presented HybridKV, a training-free KV cache compression method that combines a three-segment cache structure with layer-adaptive strategy selection. On PG-19, HybridKV consistently outperforms both StreamingLLM and SnapKV, approaching the dense baseline. Ablation studies show that the primary source of improvement is the complementary combination of recency-based and voting-based selection, while the optimal layer assignment is model-scale dependent---a finding that motivates future investigation on larger architectures.

\textbf{Code.} \url{https://github.com/lzhshs/HybridKV}. Individual reproductions: SnapKV (\url{https://github.com/lzhshs/snapkv-reproduce}), StreamingLLM (\url{https://github.com/sijiechen647-lgtm/streaming-llm-reproduction}).

\textbf{Division of work.} 刘泽辉: SnapKV reproduction, HybridKV design and implementation, experiment pipeline, paper writing. 陈思洁: StreamingLLM reproduction and integration, related work literature survey.

\textbf{AI usage disclosure.} During the research process, AI tools were used to assist with code debugging and implementation troubleshooting. All algorithm design, experimental analysis, and scientific conclusions are the authors' original work.

\section*{References}

{\small

[1] Vaswani, A., Shazeer, N., Parmar, N., et al. (2017). Attention is all you need. In \textit{NeurIPS 30}.

[2] Xiao, G., Tian, Y., Chen, B., Han, S., \& Lewis, M. (2024). Efficient streaming language models with attention sinks. In \textit{ICLR}.

[3] Li, Y., He, Y., Sun, S., Yuan, C., Hao, H., \& Tan, Y. (2024). SnapKV: LLM knows what you are looking for before generation. In \textit{NeurIPS 37}.

[4] Wu, Y., Li, Y., Xu, Z., et al. (2025). PyramidKV: Dynamic KV cache compression based on pyramidal information funneling. \textit{TMLR}.

[5] Biderman, S., Schoelkopf, H., Anthony, Q., et al. (2023). Pythia: A suite for analyzing large language models across training and scaling. In \textit{ICML}.

[6] Merity, S., Xiong, C., Bradbury, J., \& Socher, R. (2017). Pointer sentinel mixture models. In \textit{ICLR}.

[7] Rae, J.W., Potapenko, A., Jayakumar, S.M., Hillier, C., \& Lillicrap, T.P. (2020). Compressive transformers for long-range sequence modelling. In \textit{ICLR}.

[8] Sun, Y., et al. (2024). You only cache once: Decoder-decoder architectures for language models. In \textit{NeurIPS 37}.

[9] Zhang, Z., et al. (2024). H2O: Heavy-hitter oracle for efficient generative inference of large language models. In \textit{NeurIPS 36}.

[10] Dao, T., Fu, D.Y., Ermon, S., Rudra, A., \& R\'{e}, C. (2022). FlashAttention: Fast and memory-efficient exact attention with IO-awareness. In \textit{NeurIPS 35}.

}

\end{document}
